---
title: An R Markdown document converted from "WUR_historic_barley_data_analysis.ipynb"
output: html_document
---

# WUR historical phenotypic barley data analysis
origin and version of the input file: https://urgi.versailles.inrae.fr/fairdom/data_files/15?version=6

## Load packages, custom functions and data

```{r}
library(tidyverse)
library(multtest)
library(asreml)

source("../custom_functions.r")
```

```{r}
data_total <- preprocess("WR-Phenotypic-Historical-Barley-data.xlsx")
write_csv(data_total, "WUR_barley.csv")
genotypes <- distinct(data_total["accenumb"]) |> rename("genotype" = "accenumb")
cat("barley data:\n", nrow(data_total), "observations\n", nrow(genotypes), "genotypes\n", nrow(distinct(data_total["campaign"])), "campaigns\n\n")
summary(data_total[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

## Subset data, remove missing values and singletons, make factors

```{r}
data_sub_HT <- data_total |> make_data_sub("days_to_heading", 50, 250)
data_sub_PH <- data_total |> make_data_sub("plant_height", 5, 250)
data_sub_TKW <- data_total |> make_data_sub("thousand_kernel_weight", 5, 100)
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_HT |> visualise_trait(geom_density())
data_sub_PH |> visualise_trait(geom_density())
data_sub_TKW |> visualise_trait(geom_density())
```

We can see that there are not enough data points for a statistical analysis of the HT and TKW traits 

## Extract and format the campaign effects

```{r}
asreml_campaign_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PH)
```

```{r}
campaign_effect_PH <- get_campaign_effect(asreml_campaign_PH)
```

## Extract heterogeneous error variances

```{r}
error_var_PH <- get_error_var(asreml_campaign_PH)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_PH <- make_CV_df(campaign_effect_PH, error_var_PH)
```

## Remove outlier campaigns

```{r}
data_corrected_PH_I <- data_sub_PH |> correct_I(CV_df_PH)
```

## Residuals

```{r}
asreml_res_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PH_I)
```

```{r}
result_BH_PH <- make_BH_multtest(asreml_res_PH)
```

```{r}
data_corrected_PH_II <- data_corrected_PH_I |> correct_II(result_BH_PH)
```

```{r}
data_corrected_all <- data_corrected_PH_II |> pivot_longer(cols = !c(accenumb,campaign), names_to = "trait") |>
         drop_na() |> arrange(campaign, accenumb, trait)
write_csv(data_corrected_all, "WUR_barley_corrected.csv")
```

## Calculate heritability

```{r}
asreml_res_h_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PH_II)
```

```{r}
quality_PH <- get_quality(data_sub_PH, asreml_res_h_PH)
quality_all <- quality_PH |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_all,"WUR_barley_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PH_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_PH <- get_BLUEs(asreml_data_PH)

BLUE_all <- genotypes |> left_join(BLUEs_PH, by="genotype") |> 
            arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_all,"WUR_barley_BLUE_values.csv")
summary(BLUE_all)
```

```{r}
library(rmarkdown)
convert_ipynb("WUR_historic_barley_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

