---
title: An R Markdown document converted from "NARDI_historic_wheat_data_analysis.ipynb"
output: html_document
---

# NARDI historical phenotypic wheat data analysis
origin and version of the input file: https://urgi.versailles.inrae.fr/fairdom/data_files/413?version=9

## Load packages, custom functions and data

```{r}
library(tidyverse)
library(multtest)
library(asreml)

source("../custom_functions.r")
```

```{r}
data_total <- preprocess("NARDI_Phenotypic_Historical_Winter_Wheat.xlsx")
write_csv(data_total, "NARDI_wheat.csv")
genotypes <- distinct(data_total["accenumb"]) |> rename("genotype" = "accenumb")
cat("Wheat data:\n", nrow(data_total), "observations\n", nrow(genotypes), "genotypes\n", nrow(distinct(data_total["campaign"])), "campaigns\n\n")
summary(data_total[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

## Subset data, remove missing values and singletons, make factors

```{r}
data_sub_HT <- data_total |> make_data_sub("days_to_heading", 50, 250)
data_sub_PH <- data_total |> make_data_sub("plant_height", 5, 250)
data_sub_PD <- data_total |> make_data_sub("plant_density")
data_sub_TKW <- data_total |> make_data_sub("thousand_kernel_weight", 5, 100)
data_sub_HW <- data_total |> make_data_sub("hectolitric_weight")
data_sub_DMY <- data_total |> make_data_sub("dry_matter_yield")
data_sub_L <- data_total |> make_data_sub("lodging")
data_sub_LR <- data_total |> make_data_sub("leaf_rust")
data_sub_SR <- data_total |> make_data_sub("stripe_rust")
data_sub_ST <- data_total |> make_data_sub("septoria")
data_sub_PM <- data_total |> make_data_sub("powdery_mildew")
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_HT |> visualise_trait(geom_density())
data_sub_PH |> visualise_trait(geom_density())
data_sub_PD |> visualise_trait(geom_density())
data_sub_TKW |> visualise_trait(geom_density())
data_sub_HW |> visualise_trait(geom_density())
data_sub_DMY |> visualise_trait(geom_density())
data_sub_L |> visualise_trait(geom_histogram(bins=10))
data_sub_LR |> visualise_trait(geom_histogram(bins=10))
data_sub_SR |> visualise_trait(geom_histogram(bins=10))
data_sub_ST |> visualise_trait(geom_histogram(bins=10))
data_sub_PM |> visualise_trait(geom_histogram(bins=10))
```

## Extract and format the campaign effects

```{r}
asreml_campaign_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_HT)
```

```{r}
asreml_campaign_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PH)
```

```{r}
asreml_campaign_PD <- asreml(fixed = plant_density ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PD)
```

```{r}
asreml_campaign_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_TKW)
```

```{r}
asreml_campaign_HW <- asreml(fixed = hectolitric_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_HW)
```

```{r}
asreml_campaign_DMY <- asreml(fixed = dry_matter_yield ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_DMY)
```

```{r}
asreml_campaign_PM <- asreml(fixed = powdery_mildew ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_PM)
```

```{r}
campaign_effect_HT <- get_campaign_effect(asreml_campaign_HT)
campaign_effect_PH <- get_campaign_effect(asreml_campaign_PH)
campaign_effect_PD <- get_campaign_effect(asreml_campaign_PD)
campaign_effect_TKW <- get_campaign_effect(asreml_campaign_TKW)
campaign_effect_HW <- get_campaign_effect(asreml_campaign_HW)
campaign_effect_DMY <- get_campaign_effect(asreml_campaign_DMY)
campaign_effect_PM <- get_campaign_effect(asreml_campaign_PM)
```

## Extract heterogeneous error variances

```{r}
error_var_HT <- get_error_var(asreml_campaign_HT)
error_var_PH <- get_error_var(asreml_campaign_PH)
error_var_PD <- get_error_var(asreml_campaign_PD)
error_var_TKW <- get_error_var(asreml_campaign_TKW)
error_var_HW <- get_error_var(asreml_campaign_HW)
error_var_DMY <- get_error_var(asreml_campaign_DMY)
error_var_PM <- get_error_var(asreml_campaign_PM)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_HT <- make_CV_df(campaign_effect_HT, error_var_HT)
CV_df_PH <- make_CV_df(campaign_effect_PH, error_var_PH)
CV_df_PD <- make_CV_df(campaign_effect_PD, error_var_PD)
CV_df_TKW <- make_CV_df(campaign_effect_TKW, error_var_TKW)
CV_df_HW <- make_CV_df(campaign_effect_HW, error_var_HW)
CV_df_DMY <- make_CV_df(campaign_effect_DMY, error_var_DMY)
CV_df_PM <- make_CV_df(campaign_effect_PM, error_var_PM)
```

## Remove outlier campaigns

```{r}
data_corrected_HT_I <- data_sub_HT |> correct_I(CV_df_HT)
data_corrected_PH_I <- data_sub_PH |> correct_I(CV_df_PH)
data_corrected_PD_I <- data_sub_PD |> correct_I(CV_df_PD)
data_corrected_TKW_I <- data_sub_TKW |> correct_I(CV_df_TKW)
data_corrected_HW_I <- data_sub_HW |> correct_I(CV_df_HW)
data_corrected_DMY_I <- data_sub_DMY |> correct_I(CV_df_DMY)
data_corrected_PM_I <- data_sub_PM |> correct_I(CV_df_PM)
```

## Residuals

```{r}
asreml_res_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_HT_I)
```

```{r}
asreml_res_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PH_I)
```

```{r}
asreml_res_PD <- asreml(fixed = plant_density ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PD_I)
```

```{r}
asreml_res_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_TKW_I)
```

```{r}
asreml_res_DMY <- asreml(fixed = dry_matter_yield ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_DMY_I)
```

```{r}
asreml_res_PM <- asreml(fixed = powdery_mildew ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_PM_I)
```

```{r}
result_BH_HT <- make_BH_multtest(asreml_res_HT)
```

```{r}
result_BH_PH <- make_BH_multtest(asreml_res_PH)
```

```{r}
result_BH_PD <- make_BH_multtest(asreml_res_PD)
```

```{r}
result_BH_TKW <- make_BH_multtest(asreml_res_TKW)
```

```{r}
result_BH_DMY <- make_BH_multtest(asreml_res_DMY)
```

```{r}
result_BH_PM <- make_BH_multtest(asreml_res_PM)
```

```{r}
data_corrected_HT_II <- data_corrected_HT_I |> correct_II(result_BH_HT)
data_corrected_PH_II <- data_corrected_PH_I |> correct_II(result_BH_PH)
data_corrected_PD_II <- data_corrected_PD_I |> correct_II(result_BH_PD)
data_corrected_TKW_II <- data_corrected_TKW_I |> correct_II(result_BH_TKW)
data_corrected_DMY_II <- data_corrected_DMY_I |> correct_II(result_BH_DMY)
data_corrected_PM_II <- data_corrected_PM_I |> correct_II(result_BH_PM)
```

## Calculate heritability

```{r}
asreml_res_h_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_HT_II)
```

```{r}
asreml_res_h_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PH_II)
```

```{r}
asreml_res_h_PD <- asreml(plant_density ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PD_II)
```

```{r}
asreml_res_h_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_TKW_II)
```

```{r}
asreml_res_h_DMY <- asreml(dry_matter_yield ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_DMY_II)
```

```{r}
asreml_res_h_PM <- asreml(powdery_mildew ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PM_II)
```

```{r}
quality_HT <- get_quality(data_sub_HT, asreml_res_h_HT)
quality_PH <- get_quality(data_sub_PH, asreml_res_h_PH)
quality_PD <- get_quality(data_sub_PD, asreml_res_h_PD)
quality_TKW <- get_quality(data_sub_TKW, asreml_res_h_TKW)
quality_DMY <- get_quality(data_sub_DMY, asreml_res_h_DMY)
quality_PM <- get_quality(data_sub_PM, asreml_res_h_PM)

quality_all = bind_rows(quality_HT, quality_PH, quality_PD, quality_TKW, quality_DMY, quality_PM) |> 
                    mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_all,"NARDI_wheat_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_HT_II)
```

```{r}
asreml_data_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PH_II)
```

```{r}
asreml_data_PD <- asreml(plant_density ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PD_II)
```

```{r}
asreml_data_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_TKW_II)
```

```{r}
asreml_data_DMY <- asreml(dry_matter_yield ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_DMY_II)
```

```{r}
asreml_data_PM <- asreml(powdery_mildew ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PM_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_HT <- get_BLUEs(asreml_data_HT)
BLUEs_PH <- get_BLUEs(asreml_data_PH)
BLUEs_PD <- get_BLUEs(asreml_data_PD)
BLUEs_TKW <- get_BLUEs(asreml_data_TKW)
BLUEs_DMY <- get_BLUEs(asreml_data_DMY)
BLUEs_PM <- get_BLUEs(asreml_data_PM)


BLUE_all <- genotypes |> left_join(BLUEs_HT, by="genotype") |> 
                        left_join(BLUEs_PH, by="genotype") |> 
                        left_join(BLUEs_PD, by="genotype") |> 
                        left_join(BLUEs_TKW, by="genotype") |> 
                        left_join(BLUEs_DMY, by="genotype") |> 
                        left_join(BLUEs_PM, by="genotype") |>
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_all,"NARDI_wheat_BLUE_values.csv")
summary(BLUE_all)
```

```{r}
library(rmarkdown)
convert_ipynb("NARDI_historic_wheat_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

