---
title: An R Markdown document converted from "IPGR-Sadovo_historic_wheat_data_analysis.ipynb"
output: html_document
---

# IPGR-Sadovo historical phenotypic wheat data analysis
origin and version of the input file: https://urgi.versailles.inrae.fr/fairdom/data_files/14?version=6

## Load packages, custom functions and data

```{r}
library(tidyverse)
library(multtest)
library(asreml)

source("../custom_functions.r")
```

```{r}
data_total <- preprocess("20230206_123002_adjusted_IPGR-Sadovo_Historical_phenotypic_data_wheat.xlsx")
write_csv(data_total, "IPGR-Sadovo_wheat.csv")
genotypes <- distinct(data_total["accenumb"]) |> rename("genotype" = "accenumb")
cat("Wheat data:\n", nrow(data_total), "observations\n", nrow(genotypes), "genotypes\n", nrow(distinct(data_total["campaign"])), "campaigns\n\n")
summary(data_total[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

## Subset data, remove missing values and singletons, make factors

```{r}
data_sub_HT <- data_total |> make_data_sub("days_to_heading", 50, 250)
data_sub_PH <- data_total |> make_data_sub("plant_height", 5, 250)
data_sub_TKW <- data_total |> make_data_sub("thousand_kernel_weight", 5, 100)
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_HT |> visualise_trait(geom_density())
data_sub_PH |> visualise_trait(geom_density())
data_sub_TKW |> visualise_trait(geom_density())
```

## Extract and format the campaign effects

```{r}
asreml_campaign_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_HT)
```

```{r}
asreml_campaign_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PH)
```

```{r}
asreml_campaign_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_TKW)
```

```{r}
campaign_effect_HT <- get_campaign_effect(asreml_campaign_HT)
campaign_effect_PH <- get_campaign_effect(asreml_campaign_PH)
campaign_effect_TKW <- get_campaign_effect(asreml_campaign_TKW)
```

## Extract heterogeneous error variances

```{r}
error_var_HT <- get_error_var(asreml_campaign_HT)
error_var_PH <- get_error_var(asreml_campaign_PH)
error_var_TKW <- get_error_var(asreml_campaign_TKW)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_HT <- make_CV_df(campaign_effect_HT, error_var_HT)
CV_df_PH <- make_CV_df(campaign_effect_PH, error_var_PH)
CV_df_TKW <- make_CV_df(campaign_effect_TKW, error_var_TKW)
```

## Remove outlier campaigns

```{r}
data_corrected_HT_I <- data_sub_HT |> correct_I(CV_df_HT)
data_corrected_PH_I <- data_sub_PH |> correct_I(CV_df_PH)
data_corrected_TKW_I <- data_sub_TKW |> correct_I(CV_df_TKW)
```

## Residuals

```{r}
asreml_res_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_HT_I)
```

```{r}
asreml_res_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PH_I)
```

```{r}
asreml_res_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_TKW_I)
```

```{r}
result_BH_HT <- make_BH_multtest(asreml_res_HT)
```

```{r}
result_BH_PH <- make_BH_multtest(asreml_res_PH)
```

```{r}
result_BH_TKW <- make_BH_multtest(asreml_res_TKW)
```

```{r}
data_corrected_HT_II <- data_corrected_HT_I |> correct_II(result_BH_HT)
data_corrected_PH_II <- data_corrected_PH_I |> correct_II(result_BH_PH)
data_corrected_TKW_II <- data_corrected_TKW_I |> correct_II(result_BH_TKW)
```

```{r}
data_corrected_all <- bind_rows(
    data_corrected_HT_II,
    data_corrected_PH_II,
    data_corrected_TKW_II,
    ) |> pivot_longer(cols = !c(accenumb,campaign), names_to = "trait") |>
         drop_na() |> arrange(campaign, accenumb, trait)
write_csv(data_corrected_all, "IPGR-Sadovo_wheat_corrected.csv")
```

## Calculate heritability

```{r}
asreml_res_h_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_HT_II)
```

```{r}
asreml_res_h_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PH_II)
```

```{r}
asreml_res_h_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_TKW_II)
```

```{r}
quality_HT <- get_quality(data_sub_HT, asreml_res_h_HT)
quality_PH <- get_quality(data_sub_PH, asreml_res_h_PH)
quality_TKW <- get_quality(data_sub_TKW, asreml_res_h_TKW)
quality_all <- bind_rows(quality_HT, quality_PH, quality_TKW) |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_all,"IPGR-Sadovo_wheat_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_HT_II)
```

```{r}
asreml_data_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PH_II)
```

```{r}
asreml_data_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_TKW_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_HT <- get_BLUEs(asreml_data_HT)
BLUEs_PH <- get_BLUEs(asreml_data_PH)
BLUEs_TKW <- get_BLUEs(asreml_data_TKW)

BLUE_all <- genotypes |> 
                        left_join(BLUEs_HT, by="genotype") |> 
                        left_join(BLUEs_PH, by="genotype") |> 
                        left_join(BLUEs_TKW, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_all,"IPGR-Sadovo_wheat_BLUE_values.csv")
summary(BLUE_all)
```

```{r}
library(rmarkdown)
convert_ipynb("IPGR-Sadovo_historic_wheat_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

