---
title: An R Markdown document converted from "CREA-CI_historic_wheat_data_analysis.ipynb"
output: html_document
---

# CREA-CI historical phenotypic wheat data analysis

origin and version of the input file: https://urgi.versailles.inrae.fr/fairdom/data_files/196?version=10

## Load packages, custom functions and data

```{r}
library(tidyverse)
library(multtest)
library(asreml)

source("../custom_functions.r")
```

```{r}
data_total <- preprocess("CREA-historic.xlsx")
write_csv(data_total, "CREA_wheat.csv")
genotypes <- distinct(data_total["accenumb"]) |> rename("genotype" = "accenumb")
cat("Wheat data:\n", nrow(data_total), "observations\n", nrow(genotypes), "genotypes\n", nrow(distinct(data_total["campaign"])), "campaigns\n\n")
summary(data_total[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

## Subset data, remove missing values and singletons, make factors

```{r}
data_sub_HT <- data_total |> make_data_sub("days_to_heading", 50, 250)
data_sub_PH <- data_total |> make_data_sub("plant_height", 5, 250)
data_sub_TKW <- data_total |> make_data_sub("thousand_kernel_weight", 5, 100)
data_sub_PM <- data_total |> make_data_sub("powdery_mildew")
data_sub_LR<- data_total |> make_data_sub("leaf_rust")
# YR has observations in only one campaign so cannot be analysed here
data_sub_ST <- data_total |> make_data_sub("septoria_tritici_ssp")
data_sub_PC <- data_total |> make_data_sub("protein_content")
data_sub_SV <- data_total |> make_data_sub("sds_sedimentation_volume")
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_HT |> visualise_trait(geom_density())
data_sub_PH |> visualise_trait(geom_density())
data_sub_TKW |> visualise_trait(geom_density())
data_sub_PM |> visualise_trait(geom_histogram(bins=10))
data_sub_LR |> visualise_trait(geom_histogram(bins=10))
data_sub_ST |> visualise_trait(geom_histogram(bins=10))
data_sub_PC |> visualise_trait(geom_density())
data_sub_SV |> visualise_trait(geom_density())
```

## Extract and format the campaign effects

```{r}
asreml_campaign_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_HT)
```

```{r}
asreml_campaign_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PH)
```

```{r}
asreml_campaign_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_TKW)
```

```{r}
data_sub_PM <- data_sub_PM |> filter(campaign != "2013-2014") # debug asreml: this campaign has all PM values equal to zero and is excluded to enable fitting the model. 
asreml_campaign_PM <- asreml(fixed = powdery_mildew ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_PM)
```

```{r}
asreml_campaign_LR <- asreml(fixed = leaf_rust ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_LR)
```

```{r}
asreml_campaign_ST <- asreml(fixed = septoria_tritici_ssp ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_ST)
```

```{r}
asreml_campaign_PC <- asreml(fixed = protein_content ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_PC)
```

```{r}
asreml_campaign_SV <- asreml(fixed = sds_sedimentation_volume ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_SV)
```

```{r}
campaign_effect_HT <- get_campaign_effect(asreml_campaign_HT)
campaign_effect_PH <- get_campaign_effect(asreml_campaign_PH)
campaign_effect_TKW <- get_campaign_effect(asreml_campaign_TKW)
campaign_effect_PM <- get_campaign_effect(asreml_campaign_PM)
campaign_effect_LR <- get_campaign_effect(asreml_campaign_LR)
campaign_effect_ST <- get_campaign_effect(asreml_campaign_ST)
campaign_effect_PC <- get_campaign_effect(asreml_campaign_PC)
campaign_effect_SV <- get_campaign_effect(asreml_campaign_SV)
```

## Extract heterogeneous error variances

```{r}
error_var_HT <- get_error_var(asreml_campaign_HT)
error_var_PH <- get_error_var(asreml_campaign_PH)
error_var_TKW <- get_error_var(asreml_campaign_TKW)
error_var_PM <- get_error_var(asreml_campaign_PM)
error_var_LR <- get_error_var(asreml_campaign_LR)
error_var_ST <- get_error_var(asreml_campaign_ST)
error_var_PC <- get_error_var(asreml_campaign_PC)
error_var_SV <- get_error_var(asreml_campaign_SV)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_HT <- make_CV_df(campaign_effect_HT, error_var_HT)
CV_df_PH <- make_CV_df(campaign_effect_PH, error_var_PH)
CV_df_TKW <- make_CV_df(campaign_effect_TKW, error_var_TKW)
CV_df_PM <- make_CV_df(campaign_effect_PM, error_var_PM)
CV_df_LR <- make_CV_df(campaign_effect_LR, error_var_LR)
CV_df_ST <- make_CV_df(campaign_effect_ST, error_var_ST)
CV_df_PC <- make_CV_df(campaign_effect_PC, error_var_PC)
CV_df_SV <- make_CV_df(campaign_effect_SV, error_var_SV)
```

## Remove outlier campaigns

```{r}
data_corrected_HT_I <- data_sub_HT |> correct_I(CV_df_HT)
data_corrected_PH_I <- data_sub_PH |> correct_I(CV_df_PH)
data_corrected_TKW_I <- data_sub_TKW |> correct_I(CV_df_TKW)
data_corrected_PM_I <- data_sub_PM |> correct_I(CV_df_PM)
data_corrected_LR_I <- data_sub_LR |> correct_I(CV_df_LR)
data_corrected_ST_I <- data_sub_ST |> correct_I(CV_df_ST)
data_corrected_PC_I <- data_sub_PC |> correct_I(CV_df_PC)
data_corrected_SV_I <- data_sub_SV |> correct_I(CV_df_SV)
```

## Residuals

```{r}
asreml_res_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_HT_I)
```

```{r}
asreml_res_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PH_I)
```

```{r}
asreml_res_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_TKW_I)
```

```{r}
asreml_res_PM <- asreml(fixed = powdery_mildew ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_PM_I)
```

```{r}
asreml_res_LR <- asreml(fixed = leaf_rust ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_LR_I)
```

```{r}
asreml_res_ST <- asreml(fixed = septoria_tritici_ssp ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_ST_I)
```

```{r}
asreml_res_PC <- asreml(fixed = protein_content ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_PC_I)
```

```{r}
asreml_res_SV <- asreml(fixed = sds_sedimentation_volume ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SV_I)
```

```{r}
result_BH_HT <- make_BH_multtest(asreml_res_HT)
```

```{r}
result_BH_PH <- make_BH_multtest(asreml_res_PH)
```

```{r}
result_BH_TKW <- make_BH_multtest(asreml_res_TKW)
```

```{r}
result_BH_PM <- make_BH_multtest(asreml_res_PM)
```

```{r}
result_BH_LR <- make_BH_multtest(asreml_res_LR)
```

```{r}
result_BH_ST <- make_BH_multtest(asreml_res_ST)
```

```{r}
result_BH_PC <- make_BH_multtest(asreml_res_PC)
```

```{r}
result_BH_SV <- make_BH_multtest(asreml_res_SV)
```

```{r}
data_corrected_HT_II <- data_corrected_HT_I |> correct_II(result_BH_HT)
data_corrected_PH_II <- data_corrected_PH_I |> correct_II(result_BH_PH)
data_corrected_TKW_II <- data_corrected_TKW_I |> correct_II(result_BH_TKW)
data_corrected_PM_II <- data_corrected_PM_I |> correct_II(result_BH_PM)
data_corrected_LR_II <- data_corrected_LR_I |> correct_II(result_BH_LR)
data_corrected_ST_II <- data_corrected_ST_I |> correct_II(result_BH_ST)
data_corrected_PC_II <- data_corrected_PC_I |> correct_II(result_BH_PC)
data_corrected_SV_II <- data_corrected_SV_I |> correct_II(result_BH_SV)
```

```{r}
data_corrected_all <- bind_rows(
    data_corrected_HT_II,
    data_corrected_PH_II,
    data_corrected_TKW_II,
    data_corrected_PM_II,
    data_corrected_LR_II,
    data_corrected_ST_II,
    data_corrected_PC_II,
    data_corrected_SV_II
    ) |> pivot_longer(cols = !c(accenumb,campaign), names_to = "trait") |>
         drop_na() |> arrange(campaign, accenumb, trait)
write_csv(data_corrected_all, "CREA_wheat_corrected.csv")
```

## Calculate heritability

```{r}
asreml_res_h_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_HT_II)
```

```{r}
asreml_res_h_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PH_II)
```

```{r}
asreml_res_h_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_TKW_II)
```

```{r}
asreml_res_h_PM <- asreml(powdery_mildew ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PM_II)
```

```{r}
asreml_res_h_LR <- asreml(leaf_rust ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_LR_II)
```

```{r}
asreml_res_h_ST <- asreml(septoria_tritici_ssp ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_ST_II)
```

```{r}
asreml_res_h_PC <- asreml(protein_content ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PC_II)
```

```{r}
asreml_res_h_SV <- asreml(sds_sedimentation_volume ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_SV_II)
```

```{r}
quality_HT <- get_quality(data_sub_HT, asreml_res_h_HT)
quality_PH <- get_quality(data_sub_PH, asreml_res_h_PH)
quality_TKW <- get_quality(data_sub_TKW, asreml_res_h_TKW)
quality_PM <- get_quality(data_sub_PM, asreml_res_h_PM)
quality_LR <- get_quality(data_sub_LR, asreml_res_h_LR)
quality_ST <- get_quality(data_sub_ST, asreml_res_h_ST)
quality_PC <- get_quality(data_sub_PC, asreml_res_h_PC)
quality_SV <- get_quality(data_sub_SV, asreml_res_h_SV)

quality_all = bind_rows(quality_HT, quality_PH, quality_TKW, quality_PM, quality_LR, quality_ST, quality_PC, quality_SV) |> 
                    mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_all,"CREA_wheat_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_HT_II)
```

```{r}
asreml_data_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PH_II)
```

```{r}
asreml_data_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_TKW_II)
```

```{r}
asreml_data_PM <- asreml(powdery_mildew ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PM_II)
```

```{r}
asreml_data_LR <- asreml(leaf_rust ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_LR_II)
```

```{r}
asreml_data_ST <- asreml(septoria_tritici_ssp ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_ST_II)
```

```{r}
asreml_data_PC <- asreml(protein_content ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PC_II)
```

```{r}
asreml_data_SV <- asreml(sds_sedimentation_volume ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_SV_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_HT <- get_BLUEs(asreml_data_HT)
BLUEs_PH <- get_BLUEs(asreml_data_PH)
BLUEs_TKW <- get_BLUEs(asreml_data_TKW)
BLUEs_PM <- get_BLUEs(asreml_data_PM)
BLUEs_LR <- get_BLUEs(asreml_data_LR)
BLUEs_ST <- get_BLUEs(asreml_data_ST)
BLUEs_PC <- get_BLUEs(asreml_data_PC)
BLUEs_SV <- get_BLUEs(asreml_data_SV)

BLUE_all <- genotypes |> left_join(BLUEs_HT, by="genotype") |> 
                        left_join(BLUEs_PH, by="genotype") |> 
                        left_join(BLUEs_TKW, by="genotype") |> 
                        left_join(BLUEs_PM, by="genotype") |>
                        left_join(BLUEs_LR, by="genotype") |> 
                        left_join(BLUEs_ST, by="genotype") |> 
                        left_join(BLUEs_PC, by="genotype") |> 
                        left_join(BLUEs_SV, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_all,"CREA_wheat_BLUE_values.csv")
summary(BLUE_all)
```

```{r}
library(rmarkdown)
convert_ipynb("CREA-CI_historic_wheat_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

