---
title: An R Markdown document converted from "CREA-CI_historic_wheat_data_analysis.ipynb"
output: html_document
---

# CREA-CI historical phenotypic wheat data analysis

## Load packages and data

```{r}
library(readxl)
library(tidyverse)
library(multtest)
library(asreml)
```

```{r}
data_file <- "CREA-historic.xlsx"
experiments <- read_xlsx(data_file, sheet = "Experiment") |> rename_with(~tolower(gsub(" ", "_", .x)))
observations <- read_xlsx(data_file, sheet = "Observed scores") |> rename_with(~tolower(gsub(" ", "_", .x)))
glimpse(observations)
```

```{r}
CREA_wheat <- observations |>
                left_join(experiments, by = "experiment_id") |>
                unite("campaign", c(year_start,year_end), sep="-") |>
                select(accenumb, campaign, days_to_heading:sds_sedimentation_volume) |>
                drop_na(accenumb, campaign) |> distinct() |> arrange(campaign, accenumb)
write_csv(CREA_wheat, "CREA_wheat.csv")
glimpse(CREA_wheat)
```

```{r}
data_total <- CREA_wheat
genotypes <- distinct(data_total["accenumb"]) |> rename("genotype" = "accenumb")
cat(sprintf("%i observations \n%i genotypes \n%i campaigns \n\nSummary of AGENT traits values:",
            nrow(data_total), nrow(genotypes), nrow(distinct(data_total["campaign"]))))
summary(data_total[3:11])
```

## Subset data, remove missing values and singletons, make factors

```{r}
make_data_sub <- function(full_df, trait){
                    data_sub <- full_df |> select("accenumb", "campaign", all_of(trait)) |> drop_na() |> distinct() |>
                    group_by(accenumb) |> filter(n()>1) |> group_by(campaign) |> filter(n()>1) |> 
                    mutate_at(vars(accenumb, campaign), factor)
                    cat(nrow(data_sub), trait, "rows kept\n") 
                    return(as.data.frame(data_sub))}

data_sub_HT <- data_total |> make_data_sub("days_to_heading")
data_sub_PH <- data_total |> make_data_sub("plant_height")
data_sub_TKW <- data_total |> make_data_sub("thousand_kernel_weight")
data_sub_PM <- data_total |> make_data_sub("powdery_mildew")
data_sub_LR<- data_total |> make_data_sub("leaf_rust")
# YR has observations in only one campaign so cannot be analysed here
data_sub_ST <- data_total |> make_data_sub("septoria_tritici_ssp")
data_sub_PC <- data_total |> make_data_sub("protein_content")
data_sub_SV <- data_total |> make_data_sub("sds_sedimentation_volume")
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
visualise_trait <- function(data_sub, geom){
                    trait <- colnames(data_sub)[3]
                    trait_title <- trait |> str_replace_all("_", " ") |> str_to_title()
                    plot(ggplot(data_sub, aes(campaign,.data[[trait]])) + geom_jitter(na.rm=TRUE)
                         + theme(axis.text.x=element_text(angle=90,vjust=0.5,hjust=0.5))
                         + labs(title = trait_title, subtitle = "Trait values per campaign",))
                    plot(ggplot(data_sub, aes(.data[[trait]])) + geom(na.rm=TRUE)
                         + labs(title = trait_title, subtitle = "Distribution of trait values",))}
```

```{r}
data_sub_HT |> visualise_trait(geom_density)
data_sub_PH |> visualise_trait(geom_density)
data_sub_TKW |> visualise_trait(geom_density)
data_sub_PM |> visualise_trait(geom_histogram)
data_sub_LR |> visualise_trait(geom_histogram)
data_sub_ST |> visualise_trait(geom_histogram)
data_sub_PC |> visualise_trait(geom_density)
data_sub_SV |> visualise_trait(geom_density)
```

## Extract and format the campaign effects

```{r}
asreml_campaign_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_HT)
```

```{r}
asreml_campaign_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_PH)
```

```{r}
asreml_campaign_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_TKW)
```

```{r}
#asreml_campaign_PM <- asreml(fixed = powdery_mildew ~campaign, random = ~accenumb,
#                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
#                              data = data_sub_PM)
# fails with error: Singularities in the average information matrix.
```

```{r}
asreml_campaign_LR <- asreml(fixed = leaf_rust ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_LR)
```

```{r}
asreml_campaign_ST <- asreml(fixed = septoria_tritici_ssp ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_ST)
```

```{r}
asreml_campaign_PC <- asreml(fixed = protein_content ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_PC)
```

```{r}
asreml_campaign_SV <- asreml(fixed = sds_sedimentation_volume ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_SV)
```

```{r}
get_campaign_effect <- function(asreml_campaign){
                        intercept_index <- nrow(asreml_campaign$coeff$fixed)
                        intercept <- asreml_campaign$coeff$fixed[intercept_index]
                        campaign_effect <- asreml_campaign$coeff$fixed[-intercept_index,] + intercept
                        return(campaign_effect |> as.data.frame() |> rownames_to_column() |> 
                                mutate(campaign = str_sub(rowname,10)) |> 
                                select(campaign, campaign_effect))}

campaign_effect_HT <- get_campaign_effect(asreml_campaign_HT)
campaign_effect_PH <- get_campaign_effect(asreml_campaign_PH)
campaign_effect_TKW <- get_campaign_effect(asreml_campaign_TKW)
campaign_effect_LR <- get_campaign_effect(asreml_campaign_LR)
campaign_effect_ST <- get_campaign_effect(asreml_campaign_ST)
campaign_effect_PC <- get_campaign_effect(asreml_campaign_PC)
campaign_effect_SV <- get_campaign_effect(asreml_campaign_SV)
```

## Extract heterogeneous error variances

```{r}
get_error_var <- function(asreml_campaign){
                    summary(asreml_campaign)$varcomp |> slice(2:n()) |> select(std.error) |> 
                    as_tibble() |> rename("error_variance"="std.error")}

error_var_HT <- get_error_var(asreml_campaign_HT)
error_var_PH <- get_error_var(asreml_campaign_PH)
error_var_TKW <- get_error_var(asreml_campaign_TKW)
error_var_LR <- get_error_var(asreml_campaign_LR)
error_var_ST <- get_error_var(asreml_campaign_ST)
error_var_PC <- get_error_var(asreml_campaign_PC)
error_var_SV <- get_error_var(asreml_campaign_SV)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
make_CV_df <- function(campaign_effect, error_var){
                tibble(campaign_effect, error_var) |> 
                mutate(coefficient_of_variation=sqrt(error_variance)/campaign_effect) |> 
                mutate(standardized_CV=scale(coefficient_of_variation))}

CV_df_HT <- make_CV_df(campaign_effect_HT, error_var_HT)
CV_df_PH <- make_CV_df(campaign_effect_PH, error_var_PH)
CV_df_TKW <- make_CV_df(campaign_effect_TKW, error_var_TKW)
CV_df_LR <- make_CV_df(campaign_effect_LR, error_var_LR)
CV_df_ST <- make_CV_df(campaign_effect_ST, error_var_ST)
CV_df_PC <- make_CV_df(campaign_effect_PC, error_var_PC)
CV_df_SV <- make_CV_df(campaign_effect_SV, error_var_SV)
```

## Remove outlier campaigns

```{r}
correct_I <- function(data_sub, CV_df){
                        outlier_campaigns <- CV_df |> filter(standardized_CV > 3.5) |> select(campaign)
                        cat(nrow(outlier_campaigns), "outlier campaign(s) detected for trait", colnames(data_sub)[3], "\n")
                        return(data_sub |> filter(! campaign %in% outlier_campaigns) |>
                               group_by(campaign) |> filter(n()>1) |> ungroup())}

data_corrected_HT_I <- data_sub_HT |> correct_I(CV_df_HT)
data_corrected_PH_I <- data_sub_PH |> correct_I(CV_df_PH)
data_corrected_TKW_I <- data_sub_TKW |> correct_I(CV_df_TKW)
data_corrected_LR_I <- data_sub_LR |> correct_I(CV_df_LR)
data_corrected_ST_I <- data_sub_ST |> correct_I(CV_df_ST)
data_corrected_PC_I <- data_sub_PC |> correct_I(CV_df_PC)
data_corrected_SV_I <- data_sub_SV |> correct_I(CV_df_SV)
```

## Residuals

```{r}
asreml_res_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_HT_I)
```

```{r}
asreml_res_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_PH_I)
```

```{r}
asreml_res_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_TKW_I)
```

```{r}
asreml_res_LR <- asreml(fixed = leaf_rust ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_LR_I)
```

```{r}
asreml_res_ST <- asreml(fixed = septoria_tritici_ssp ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_ST_I)
```

```{r}
asreml_res_PC <- asreml(fixed = protein_content ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_PC_I)
```

```{r}
asreml_res_SV <- asreml(fixed = sds_sedimentation_volume ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SV_I)
```

```{r}
make_BH_multtest <- function(asreml_res){
            trait_title <- colnames(asreml_res$mf)[3] |> str_replace_all("_", " ") |> str_to_title()
            residual <- asreml_res$residuals
            MAD <- 1.4826*median(abs(residual-median(residual))) #approx constant https://en.wikipedia.org/wiki/Median_absolute_deviation#Relation_to_standard_deviation
            rawp_BHStud <- 2 * (1 - pnorm(abs(residual/MAD)))
            test_BHStud <- mt.rawp2adjp(rawp_BHStud, proc=c("Holm"))
            result_df <- tibble(adjp=test_BHStud[[1]][,1], bholm=test_BHStud[[1]][,2], index=test_BHStud[[2]]) |>
                    mutate(is_outlier = bholm < 0.05) |> arrange(index) |> mutate(std_residual = scale(residual))
            cat(sum(result_df$is_outlier), "outlier value(s) detected for trait", trait_title)
            plot(ggplot(result_df, aes(index,std_residual, color=is_outlier)) + geom_jitter()
                 + scale_colour_manual(values = c("TRUE" = "#A51D2D", "FALSE" = "#26A269"))
                 + labs(title = trait_title, subtitle = "Outlier detection with Bonferroni–Holm method"))
            return(result_df |> as.data.frame())}
```

```{r}
result_BH_HT <- make_BH_multtest(asreml_res_HT)
```

```{r}
result_BH_PH <- make_BH_multtest(asreml_res_PH)
```

```{r}
result_BH_TKW <- make_BH_multtest(asreml_res_TKW)
```

```{r}
result_BH_LR <- make_BH_multtest(asreml_res_LR)
```

```{r}
result_BH_ST <- make_BH_multtest(asreml_res_ST)
```

```{r}
result_BH_PC <- make_BH_multtest(asreml_res_PC)
```

```{r}
result_BH_SV <- make_BH_multtest(asreml_res_SV)
```

```{r}
correct_II <- function(data_corrected_I, result_BH){
                        data_corrected_I[result_BH[which(!result_BH$is_outlier),"index"],] |> 
                        group_by(accenumb) |> filter(n() > 1) |> arrange(campaign) |> 
                        as.data.frame() |> droplevels()}

data_corrected_HT_II <- data_corrected_HT_I |> correct_II(result_BH_HT)
data_corrected_PH_II <- data_corrected_PH_I |> correct_II(result_BH_PH)
data_corrected_TKW_II <- data_corrected_TKW_I |> correct_II(result_BH_TKW)
data_corrected_LR_II <- data_corrected_LR_I |> correct_II(result_BH_LR)
data_corrected_ST_II <- data_corrected_ST_I |> correct_II(result_BH_ST)
data_corrected_PC_II <- data_corrected_PC_I |> correct_II(result_BH_PC)
data_corrected_SV_II <- data_corrected_SV_I |> correct_II(result_BH_SV)
```

## Calculate heritability

```{r}
asreml_res_h_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_HT_II)
```

```{r}
asreml_res_h_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PH_II)
```

```{r}
asreml_res_h_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_TKW_II)
```

```{r}
asreml_res_h_LR <- asreml(leaf_rust ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_LR_II)
```

```{r}
asreml_res_h_ST <- asreml(septoria_tritici_ssp ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_ST_II)
```

```{r}
asreml_res_h_PC <- asreml(protein_content ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_PC_II)
```

```{r}
asreml_res_h_SV <- asreml(sds_sedimentation_volume ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_SV_II)
```

```{r}
get_quality <- function(data_sub, asreml_res_h){
                trait <- colnames(data_sub)[3]
                varcomp <- summary(asreml_res_h)$varcomp
                var_G <- varcomp[2,"component"]
                var_E <- mean(varcomp[3:nrow(varcomp),"component"])
                no_campaigns <- mean(table(data_sub$accenumb[!is.na(data_sub[,trait])]))
                heritability <- var_G/(var_G+(var_E/no_campaigns))
                cat(colnames(data_sub)[3], "heritability:", round(heritability*100,1), "%\n")
                return(tibble(trait, heritability, var_G, var_E, no_campaigns))}

quality_HT <- get_quality(data_sub_HT, asreml_res_h_HT)
quality_PH <- get_quality(data_sub_PH, asreml_res_h_PH)
quality_TKW <- get_quality(data_sub_TKW, asreml_res_h_TKW)
quality_LR <- get_quality(data_sub_LR, asreml_res_h_LR)
quality_ST <- get_quality(data_sub_ST, asreml_res_h_ST)
quality_PC <- get_quality(data_sub_PC, asreml_res_h_PC)
quality_SV <- get_quality(data_sub_SV, asreml_res_h_SV)

quality_all = bind_rows(quality_HT, quality_PH, quality_TKW, quality_LR, quality_ST, quality_PC, quality_SV) |> 
                    mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_all,"CREA_wheat_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_HT_II)
```

```{r}
asreml_data_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PH_II)
```

```{r}
asreml_data_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_TKW_II)
```

```{r}
asreml_data_LR <- asreml(leaf_rust ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_LR_II)
```

```{r}
asreml_data_ST <- asreml(septoria_tritici_ssp ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_ST_II)
```

```{r}
asreml_data_PC <- asreml(protein_content ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_PC_II)
```

```{r}
asreml_data_SV <- asreml(sds_sedimentation_volume ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_SV_II)
```

## Extract BLUEs and create output file

```{r}
get_BLUEs <- function(asreml_data){
                            intercept_index <- nrow(asreml_data$coeff$fixed)
                            intercept <- asreml_data$coeff$fixed[intercept_index]
                            BLUE <- asreml_data$coeff$fixed[-intercept_index,] + intercept
                            return(BLUE |> as.data.frame() |>
                                    rownames_to_column() |>
                                    mutate(genotype=str_split_i(rowname,"_", -1)) |> #assuming there is no underscore in the accession numbers
                                    select(genotype, BLUE))}

BLUEs_HT <- get_BLUEs(asreml_data_HT) |> rename("days_to_heading_BLUE"="BLUE")
BLUEs_PH <- get_BLUEs(asreml_data_PH) |> rename("plant_height_BLUE"="BLUE")
BLUEs_TKW <- get_BLUEs(asreml_data_TKW) |> rename("thousand_kernel_weight_BLUE"="BLUE")
BLUEs_LR <- get_BLUEs(asreml_data_LR) |> rename("leaf_rust_BLUE"="BLUE")
BLUEs_ST <- get_BLUEs(asreml_data_ST) |> rename("septoria_tritici_ssp_BLUE"="BLUE")
BLUEs_PC <- get_BLUEs(asreml_data_PC) |> rename("protein_content_BLUE"="BLUE")
BLUEs_SV <- get_BLUEs(asreml_data_SV) |> rename("sds_sedimentation_volume_BLUE"="BLUE")


BLUE_all <- genotypes |> left_join(BLUEs_HT, by="genotype") |> 
                        left_join(BLUEs_PH, by="genotype") |> 
                        left_join(BLUEs_TKW, by="genotype") |> 
                        left_join(BLUEs_LR, by="genotype") |> 
                        left_join(BLUEs_ST, by="genotype") |> 
                        left_join(BLUEs_PC, by="genotype") |> 
                        left_join(BLUEs_SV, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_all,"CREA_wheat_BLUE_values.csv")
summary(BLUE_all)
```

```{r}
library(rmarkdown)
convert_ipynb("CREA-CI_historic_wheat_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

