---
title: An R Markdown document converted from "IHAR_historic_wheat_data_analysis.ipynb"
output: html_document
---

# IHAR historical phenotypic wheat data analysis
origin and version of the input files: 
- https://urgi.versailles.inrae.fr/fairdom/data_files/54?version=4
- https://urgi.versailles.inrae.fr/fairdom/data_files/55?version=3

## Load packages, custom functions and data

```{r}
library(tidyverse)
library(multtest)
library(asreml)
asreml.options(workspace="20gb")
source("../custom_functions.r")
```

```{r}
data_total_winter <- preprocess("IHAR Phenotypic_Historical-winter wheat.xlsx")
write_csv(data_total_winter, "IHAR_winter_wheat.csv")
genotypes_w <- distinct(data_total_winter["accenumb"]) |> rename("genotype" = "accenumb")
cat("Winter Wheat:\n", nrow(data_total_winter), "observations\n", nrow(genotypes_w), "genotypes\n", nrow(distinct(data_total_winter["campaign"])), "campaigns\n\n")
summary(data_total_winter[c("beginning_of_heading", "end_of_heading", "plant_height", "thousand_kernel_weight")])
```

```{r}
data_total_spring <- preprocess("IHAR Phenotypic_Historical-spring wheat.xlsx") |> 
                        mutate(beginning_of_heading = heading_date_to_days(beginning_of_heading_date)) |>
                        mutate(end_of_heading = heading_date_to_days(end_of_heading_date))
write_csv(data_total_spring, "IHAR_spring_wheat.csv")
genotypes_s <- distinct(data_total_spring["accenumb"]) |> rename("genotype" = "accenumb")
cat("Spring Wheat:\n", nrow(data_total_spring), "observations\n", nrow(genotypes_s), "genotypes\n", nrow(distinct(data_total_spring["campaign"])), "campaigns\n\n")
summary(data_total_spring[c("beginning_of_heading", "end_of_heading", "plant_height", "thousand_kernel_weight")])
```

Let's choose "end of heading" as heading time trait, as it has way less NAs than "beginning of heading"

## Subset data, remove missing values and singletons, make factors

```{r}
cat("Winter:\n")
data_sub_WW_HT <- data_total_winter |> make_data_sub("end_of_heading", 50, 250)
data_sub_WW_PH <- data_total_winter |> make_data_sub("plant_height", 5, 250)
data_sub_WW_TKW <- data_total_winter |> make_data_sub("thousand_kernel_weight", 5, 100)
```

```{r}
cat("Spring:\n")
data_sub_SW_HT <- data_total_spring |> make_data_sub("end_of_heading", 50, 250)
data_sub_SW_PH <- data_total_spring |> make_data_sub("plant_height", 5, 250)
data_sub_SW_TKW <- data_total_spring |> make_data_sub("thousand_kernel_weight", 5, 100)
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_WW_HT |> visualise_trait(geom_density())
data_sub_WW_PH |> visualise_trait(geom_density())
data_sub_WW_TKW |> visualise_trait(geom_density())
```

```{r}
data_sub_SW_HT |> visualise_trait(geom_density())
data_sub_SW_PH |> visualise_trait(geom_density())
data_sub_SW_TKW |> visualise_trait(geom_density())
```

## Extract and format the campaign effects

```{r}
asreml_campaign_WW_HT <- asreml(fixed = end_of_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_WW_HT)
```

```{r}
asreml_campaign_WW_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_WW_PH)
```

```{r}
asreml_campaign_WW_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_WW_TKW)
```

```{r}
asreml_campaign_SW_HT <- asreml(fixed = end_of_heading ~campaign, random = ~accenumb, 
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                              data = data_sub_SW_HT)
```

```{r}
asreml_campaign_SW_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_SW_PH)
```

```{r}
asreml_campaign_SW_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_SW_TKW)
```

```{r}
campaign_effect_WW_HT <- get_campaign_effect(asreml_campaign_WW_HT)
campaign_effect_WW_PH <- get_campaign_effect(asreml_campaign_WW_PH)
campaign_effect_WW_TKW <- get_campaign_effect(asreml_campaign_WW_TKW)
```

```{r}
campaign_effect_SW_HT <- get_campaign_effect(asreml_campaign_SW_HT)
campaign_effect_SW_PH <- get_campaign_effect(asreml_campaign_SW_PH)
campaign_effect_SW_TKW <- get_campaign_effect(asreml_campaign_SW_TKW)
```

## Extract heterogeneous error variances

```{r}
error_var_WW_HT <- get_error_var(asreml_campaign_WW_HT)
error_var_WW_PH <- get_error_var(asreml_campaign_WW_PH)
error_var_WW_TKW <- get_error_var(asreml_campaign_WW_TKW)
```

```{r}
error_var_SW_HT <- get_error_var(asreml_campaign_SW_HT)
error_var_SW_PH <- get_error_var(asreml_campaign_SW_PH)
error_var_SW_TKW <- get_error_var(asreml_campaign_SW_TKW)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_WW_HT <- make_CV_df(campaign_effect_WW_HT, error_var_WW_HT)
CV_df_WW_PH <- make_CV_df(campaign_effect_WW_PH, error_var_WW_PH)
CV_df_WW_TKW <- make_CV_df(campaign_effect_WW_TKW, error_var_WW_TKW)
```

```{r}
CV_df_SW_HT <- make_CV_df(campaign_effect_SW_HT, error_var_SW_HT)
CV_df_SW_PH <- make_CV_df(campaign_effect_SW_PH, error_var_SW_PH)
CV_df_SW_TKW <- make_CV_df(campaign_effect_SW_TKW, error_var_SW_TKW)
```

## Remove outlier campaigns

```{r}
data_corrected_WW_HT_I <- data_sub_WW_HT |> correct_I(CV_df_WW_HT)
data_corrected_WW_PH_I <- data_sub_WW_PH |> correct_I(CV_df_WW_PH)
data_corrected_WW_TKW_I <- data_sub_WW_TKW |> correct_I(CV_df_WW_TKW)
```

```{r}
data_corrected_SW_HT_I <- data_sub_SW_HT |> correct_I(CV_df_SW_HT)
data_corrected_SW_PH_I <- data_sub_SW_PH |> correct_I(CV_df_SW_PH)
data_corrected_SW_TKW_I <- data_sub_SW_TKW |> correct_I(CV_df_SW_TKW)
```

## Residuals

```{r}
asreml_res_WW_HT <- asreml(fixed = end_of_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_WW_HT_I)
```

```{r}
asreml_res_WW_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_WW_PH_I)
```

```{r}
asreml_res_WW_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_WW_TKW_I)
```

```{r}
asreml_res_SW_HT <- asreml(fixed = end_of_heading ~ accenumb, random = ~ campaign,
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SW_HT_I)
```

```{r}
asreml_res_SW_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_SW_PH_I)
```

```{r}
asreml_res_SW_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SW_TKW_I)
```

```{r}
result_BH_WW_HT <- make_BH_multtest(asreml_res_WW_HT)
result_BH_WW_PH <- make_BH_multtest(asreml_res_WW_PH)
result_BH_WW_TKW <- make_BH_multtest(asreml_res_WW_TKW)
```

```{r}
result_BH_SW_HT <- make_BH_multtest(asreml_res_SW_HT)
result_BH_SW_PH <- make_BH_multtest(asreml_res_SW_PH)
result_BH_SW_TKW <- make_BH_multtest(asreml_res_SW_TKW)
```

```{r}
data_corrected_WW_HT_II <- data_corrected_WW_HT_I |> correct_II(result_BH_WW_HT)
data_corrected_WW_PH_II <- data_corrected_WW_PH_I |> correct_II(result_BH_WW_PH)
data_corrected_WW_TKW_II <- data_corrected_WW_TKW_I |> correct_II(result_BH_WW_TKW)
```

```{r}
data_corrected_SW_HT_II <- data_corrected_SW_HT_I |> correct_II(result_BH_SW_HT)
data_corrected_SW_PH_II <- data_corrected_SW_PH_I |> correct_II(result_BH_SW_PH)
data_corrected_SW_TKW_II <- data_corrected_SW_TKW_I |> correct_II(result_BH_SW_TKW)
```

## Calculate heritability

```{r}
asreml_res_h_WW_HT <- asreml(end_of_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_WW_HT_II)
```

```{r}
asreml_res_h_WW_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_WW_PH_II)
```

```{r}
asreml_res_h_WW_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_WW_TKW_II)
```

```{r}
asreml_res_h_SW_HT <- asreml(end_of_heading ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"), 
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_SW_HT_II)
```

```{r}
asreml_res_h_SW_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_SW_PH_II)
```

```{r}
asreml_res_h_SW_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                     na.action = na.method(x="omit", y="omit"),
                     residual = ~dsum( ~id(units) | campaign), data = data_corrected_SW_TKW_II)
```

```{r}
quality_WW_HT <- get_quality(data_sub_WW_HT, asreml_res_h_WW_HT)
quality_WW_PH <- get_quality(data_sub_WW_PH, asreml_res_h_WW_PH)
quality_WW_TKW <- get_quality(data_sub_WW_TKW, asreml_res_h_WW_TKW)
quality_WW_all <- bind_rows(quality_WW_HT, quality_WW_PH, quality_WW_TKW) |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_WW_all,"IHAR_winter_wheat_data_quality.csv")
```

```{r}
quality_SW_HT <- get_quality(data_sub_SW_HT, asreml_res_h_SW_HT)
quality_SW_PH <- get_quality(data_sub_SW_PH, asreml_res_h_SW_PH)
quality_SW_TKW <- get_quality(data_sub_SW_TKW, asreml_res_h_SW_TKW)
quality_SW_all <- bind_rows(quality_SW_HT, quality_SW_PH, quality_SW_TKW) |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_SW_all,"IHAR_spring_wheat_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_WW_HT <- asreml(end_of_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_WW_HT_II)
```

```{r}
asreml_data_WW_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_WW_PH_II)
```

```{r}
asreml_data_WW_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                           na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                           data = data_corrected_WW_TKW_II)
```

```{r}
asreml_data_SW_HT <- asreml(end_of_heading ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_SW_HT_II)
```

```{r}
asreml_data_SW_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_SW_PH_II)
```

```{r}
asreml_data_SW_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                           na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                           data = data_corrected_SW_TKW_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_WW_HT <- get_BLUEs(asreml_data_WW_HT)
BLUEs_WW_PH <- get_BLUEs(asreml_data_WW_PH)
BLUEs_WW_TKW <- get_BLUEs(asreml_data_WW_TKW)

BLUE_WW_all <- genotypes_w |> 
                        left_join(BLUEs_WW_HT, by="genotype") |> 
                        left_join(BLUEs_WW_PH, by="genotype") |> 
                        left_join(BLUEs_WW_TKW, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_WW_all,"IHAR_winter_wheat_BLUE_values.csv")
summary(BLUE_WW_all)
```

```{r}
BLUEs_SW_HT <- get_BLUEs(asreml_data_SW_HT)
BLUEs_SW_PH <- get_BLUEs(asreml_data_SW_PH)
BLUEs_SW_TKW <- get_BLUEs(asreml_data_SW_TKW)


BLUE_SW_all <- genotypes_s |> 
                        left_join(BLUEs_SW_HT, by="genotype") |> 
                        left_join(BLUEs_SW_PH, by="genotype") |> 
                        left_join(BLUEs_SW_TKW, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_SW_all,"IHAR_spring_wheat_BLUE_values.csv")
summary(BLUE_SW_all)
```

```{r}
library(rmarkdown)
convert_ipynb("IHAR_historic_wheat_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

