---
title: An R Markdown document converted from "NPPC_historic_barley_data_analysis.ipynb"
output: html_document
---

# NPPC historical phenotypic barley data analysis

## Load packages, custom functions and data

```{r}
library(readxl)
library(tidyverse)
library(multtest)
library(asreml)

source("../custom_functions.r")
```

```{r}
data_total_winter <- preprocess("Historic_Barley_W_NPPC.xlsx")
write_csv(data_total_winter, "NPPC_winter_barley.csv")
genotypes_w <- distinct(data_total_winter["accenumb"]) |> rename("genotype" = "accenumb")
cat("Winter Barley:\n", nrow(data_total_winter), "observations\n", nrow(genotypes_w), "genotypes\n", nrow(distinct(data_total_winter["campaign"])), "campaigns\n\n")
summary(data_total_winter[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

```{r}
data_total_spring <- preprocess("Historic_Barley_S_NPPC.xlsx")
write_csv(data_total_spring, "NPPC_spring_barley.csv")
genotypes_s <- distinct(data_total_spring["accenumb"]) |> rename("genotype" = "accenumb")
cat("Spring Barley:\n", nrow(data_total_spring), "observations\n", nrow(genotypes_s), "genotypes\n", nrow(distinct(data_total_spring["campaign"])), "campaigns\n\n")
summary(data_total_spring[c("days_to_heading", "plant_height", "thousand_kernel_weight")])
```

## Subset data, remove missing values and singletons, make factors

```{r}
cat("Winter:\n")
data_sub_WB_HT <- data_total_winter |> make_data_sub("days_to_heading")
data_sub_WB_PH <- data_total_winter |> make_data_sub("plant_height")
data_sub_WB_TKW <- data_total_winter |> make_data_sub("thousand_kernel_weight")
```

```{r}
cat("Spring:\n")
data_sub_SB_HT <- data_total_spring |> make_data_sub("days_to_heading")
data_sub_SB_PH <- data_total_spring |> make_data_sub("plant_height")
data_sub_SB_TKW <- data_total_spring |> make_data_sub("thousand_kernel_weight")
```

## Visualise trait values 

```{r}
options(repr.plot.width=12, repr.plot.height=7)
data_sub_WB_HT |> visualise_trait(geom_density())
data_sub_WB_PH |> visualise_trait(geom_density())
data_sub_WB_TKW |> visualise_trait(geom_density())
```

```{r}
data_sub_SB_HT |> visualise_trait(geom_density())
data_sub_SB_PH |> visualise_trait(geom_density())
data_sub_SB_TKW |> visualise_trait(geom_density())
```

## Extract and format the campaign effects

```{r}
asreml_campaign_WB_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                             data = data_sub_WB_HT)
```

```{r}
asreml_campaign_WB_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_WB_PH)
```

```{r}
asreml_campaign_WB_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_WB_TKW)
```

```{r}
asreml_campaign_SB_HT <- asreml(fixed = days_to_heading ~campaign, random = ~accenumb, 
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign),
                              data = data_sub_SB_HT)
```

```{r}
asreml_campaign_SB_PH <- asreml(fixed = plant_height ~campaign, random = ~accenumb,
                             na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                             data = data_sub_SB_PH)
```

```{r}
asreml_campaign_SB_TKW <- asreml(fixed = thousand_kernel_weight ~campaign, random = ~accenumb,
                              na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                              data = data_sub_SB_TKW)
```

```{r}
campaign_effect_WB_HT <- get_campaign_effect(asreml_campaign_WB_HT)
campaign_effect_WB_PH <- get_campaign_effect(asreml_campaign_WB_PH)
campaign_effect_WB_TKW <- get_campaign_effect(asreml_campaign_WB_TKW)
```

```{r}
campaign_effect_SB_HT <- get_campaign_effect(asreml_campaign_SB_HT)
campaign_effect_SB_PH <- get_campaign_effect(asreml_campaign_SB_PH)
campaign_effect_SB_TKW <- get_campaign_effect(asreml_campaign_SB_TKW)
```

## Extract heterogeneous error variances

```{r}
error_var_WB_HT <- get_error_var(asreml_campaign_WB_HT)
error_var_WB_PH <- get_error_var(asreml_campaign_WB_PH)
error_var_WB_TKW <- get_error_var(asreml_campaign_WB_TKW)
```

```{r}
error_var_SB_HT <- get_error_var(asreml_campaign_SB_HT)
error_var_SB_PH <- get_error_var(asreml_campaign_SB_PH)
error_var_SB_TKW <- get_error_var(asreml_campaign_SB_TKW)
```

## Calculate and standardize the coefficient of variation (CV)

```{r}
CV_df_WB_HT <- make_CV_df(campaign_effect_WB_HT, error_var_WB_HT)
CV_df_WB_PH <- make_CV_df(campaign_effect_WB_PH, error_var_WB_PH)
CV_df_WB_TKW <- make_CV_df(campaign_effect_WB_TKW, error_var_WB_TKW)
```

```{r}
CV_df_SB_HT <- make_CV_df(campaign_effect_SB_HT, error_var_SB_HT)
CV_df_SB_PH <- make_CV_df(campaign_effect_SB_PH, error_var_SB_PH)
CV_df_SB_TKW <- make_CV_df(campaign_effect_SB_TKW, error_var_SB_TKW)
```

## Remove outlier campaigns

```{r}
data_corrected_WB_HT_I <- data_sub_WB_HT |> correct_I(CV_df_WB_HT)
data_corrected_WB_PH_I <- data_sub_WB_PH |> correct_I(CV_df_WB_PH)
data_corrected_WB_TKW_I <- data_sub_WB_TKW |> correct_I(CV_df_WB_TKW)
```

```{r}
data_corrected_SB_HT_I <- data_sub_SB_HT |> correct_I(CV_df_SB_HT)
data_corrected_SB_PH_I <- data_sub_SB_PH |> correct_I(CV_df_SB_PH)
data_corrected_SB_TKW_I <- data_sub_SB_TKW |> correct_I(CV_df_SB_TKW)
```

## Residuals

```{r}
asreml_res_WB_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_WB_HT_I)
```

```{r}
asreml_res_WB_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_WB_PH_I)
```

```{r}
asreml_res_WB_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_WB_TKW_I)
```

```{r}
asreml_res_SB_HT <- asreml(fixed = days_to_heading ~ accenumb, random = ~ campaign,
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SB_HT_I)
```

```{r}
asreml_res_SB_PH <- asreml(fixed = plant_height ~ accenumb, random = ~ campaign, 
                 na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                 data = data_corrected_SB_PH_I)
```

```{r}
asreml_res_SB_TKW <- asreml(fixed = thousand_kernel_weight ~ accenumb, random = ~ campaign, 
                  na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                  data = data_corrected_SB_TKW_I)
```

```{r}
result_BH_WB_HT <- make_BH_multtest(asreml_res_WB_HT)
result_BH_WB_PH <- make_BH_multtest(asreml_res_WB_PH)
result_BH_WB_TKW <- make_BH_multtest(asreml_res_WB_TKW)
```

```{r}
result_BH_SB_HT <- make_BH_multtest(asreml_res_SB_HT)
result_BH_SB_PH <- make_BH_multtest(asreml_res_SB_PH)
result_BH_SB_TKW <- make_BH_multtest(asreml_res_SB_TKW)
```

```{r}
data_corrected_WB_HT_II <- data_corrected_WB_HT_I |> correct_II(result_BH_WB_HT)
data_corrected_WB_PH_II <- data_corrected_WB_PH_I |> correct_II(result_BH_WB_PH)
data_corrected_WB_TKW_II <- data_corrected_WB_TKW_I |> correct_II(result_BH_WB_TKW)
```

```{r}
data_corrected_SB_HT_II <- data_corrected_SB_HT_I |> correct_II(result_BH_SB_HT)
data_corrected_SB_PH_II <- data_corrected_SB_PH_I |> correct_II(result_BH_SB_PH)
data_corrected_SB_TKW_II <- data_corrected_SB_TKW_I |> correct_II(result_BH_SB_TKW)
```

## Calculate heritability

```{r}
asreml_res_h_WB_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_WB_HT_II)
```

```{r}
asreml_res_h_WB_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_WB_PH_II)
```

```{r}
asreml_res_h_WB_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"),
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_WB_TKW_II)
```

```{r}
asreml_res_h_SB_HT <- asreml(days_to_heading ~1, random = ~accenumb + campaign, 
                    na.action = na.method(x="omit", y="omit"), 
                    residual = ~dsum( ~id(units) | campaign), data = data_corrected_SB_HT_II)
```

```{r}
asreml_res_h_SB_PH <- asreml(plant_height ~1, random = ~accenumb + campaign, 
                   na.action = na.method(x="omit", y="omit"), 
                   residual = ~dsum( ~id(units) | campaign), data = data_corrected_SB_PH_II)
```

```{r}
asreml_res_h_SB_TKW <- asreml(thousand_kernel_weight ~1, random = ~accenumb + campaign, 
                     na.action = na.method(x="omit", y="omit"),
                     residual = ~dsum( ~id(units) | campaign), data = data_corrected_SB_TKW_II)
```

```{r}
quality_WB_HT <- get_quality(data_sub_WB_HT, asreml_res_h_WB_HT)
quality_WB_PH <- get_quality(data_sub_WB_PH, asreml_res_h_WB_PH)
quality_WB_TKW <- get_quality(data_sub_WB_TKW, asreml_res_h_WB_TKW)
quality_WB_all <- bind_rows(quality_WB_HT, quality_WB_PH, quality_WB_TKW) |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_WB_all,"NPPC_winter_barley_data_quality.csv")
```

```{r}
quality_SB_HT <- get_quality(data_sub_SB_HT, asreml_res_h_SB_HT)
quality_SB_PH <- get_quality(data_sub_SB_PH, asreml_res_h_SB_PH)
quality_SB_TKW <- get_quality(data_sub_SB_TKW, asreml_res_h_SB_TKW)
quality_SB_all <- bind_rows(quality_SB_HT, quality_SB_PH, quality_SB_TKW) |> mutate(across(-1, \(x) round(x, 3)))
write_csv(quality_SB_all,"NPPC_spring_barley_data_quality.csv")
```

## Run linear mixed model with outlier corrected data

```{r}
asreml_data_WB_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_WB_HT_II)
```

```{r}
asreml_data_WB_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_WB_PH_II)
```

```{r}
asreml_data_WB_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                           na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                           data = data_corrected_WB_TKW_II)
```

```{r}
asreml_data_SB_HT <- asreml(days_to_heading ~accenumb, random = ~campaign, 
                          na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                          data = data_corrected_SB_HT_II)
```

```{r}
asreml_data_SB_PH <- asreml(plant_height ~accenumb, random = ~campaign, 
                         na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                         data = data_corrected_SB_PH_II)
```

```{r}
asreml_data_SB_TKW <- asreml(thousand_kernel_weight ~accenumb, random = ~campaign, 
                           na.action = na.method(x="omit", y="omit"), residual = ~dsum(~id(units) | campaign), 
                           data = data_corrected_SB_TKW_II)
```

## Extract BLUEs and create output file

```{r}
BLUEs_WB_HT <- get_BLUEs(asreml_data_WB_HT)
BLUEs_WB_PH <- get_BLUEs(asreml_data_WB_PH)
BLUEs_WB_TKW <- get_BLUEs(asreml_data_WB_TKW)

BLUE_WB_all <- genotypes_w |> 
                        left_join(BLUEs_WB_HT, by="genotype") |> 
                        left_join(BLUEs_WB_PH, by="genotype") |> 
                        left_join(BLUEs_WB_TKW, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_WB_all,"NPPC_winter_barley_BLUE_values.csv")
summary(BLUE_WB_all)
```

```{r}
BLUEs_SB_HT <- get_BLUEs(asreml_data_SB_HT)
BLUEs_SB_PH <- get_BLUEs(asreml_data_SB_PH)
BLUEs_SB_TKW <- get_BLUEs(asreml_data_SB_TKW)


BLUE_SB_all <- genotypes_s |> 
                        left_join(BLUEs_SB_HT, by="genotype") |> 
                        left_join(BLUEs_SB_PH, by="genotype") |> 
                        left_join(BLUEs_SB_TKW, by="genotype") |> 
                        arrange(genotype) |> mutate(across(-1, \(x) round(x, 2)))

write_csv(BLUE_SB_all,"NPPC_spring_barley_BLUE_values.csv")
summary(BLUE_SB_all)
```

```{r}
library(rmarkdown)
convert_ipynb("NPPC_historic_barley_data_analysis.ipynb")
```

```{r}
sessionInfo()
```

